Do motif roles provide unique information about species risk of extinction?

### 2021-07-20 manuscript/motifs_vs_degree.tex

% Maybe Proc B - approached Alyssa re: other motifs paper, flexible length limit.

@@

    % In the Bayesian network framework, only the stable motifs are included.
    % We show that even in this case participation in different motifs is associated with different extinction risk.
    % Participation in different motifs is also correlated with in-degree and trophic level.
    % In a Bayesian network, trophic level has a very large effect on extinction risk, and likely explains a lot of what we observe for degree.
    % However, the trends we observe for motifs are not totally explained by the correlations of motifs with trophic level and degree, indicating that both simple and meso-scale descriptions of species roles provide unique information about extinction risk.

    % Omnivory may be stabilizing at high connectances because there is a low-TL "rescue prey" available to a high-risk, high-TL species. At lower connectances, perhaps higher-TL species encounter fewer disturbances and the rescue prey is less important?
    % In Lotka-Volterra systems, where top-down effects can also cause extinction, omnivory tends to be associated with less persistence. This suggests that what happens in real systems could depend on the dominant type of regulation.

    % - Add a panel showing in-degree at different TL's, like in Anna's fig. Or switch to persistence vs. in-deg for each TL in a panel.
\clearpage
\section*{Introduction}

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad of species in food webs depend for their survival \citep{}. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{} and secondary production \citep{}, soil respiration and carbon cycling \citep{} and consumer diversity \citep{scherber2010bottom, Baiser2016}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community promotes stability across all different levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers, e.g. herbivores, predators, and parasitoids \citep{scherber2010bottom}.
     %This has been shown in both theoretical work~\citep{} and empirical experiments ~\citep{} and results from the vertical propagation of disturbances through the network of interacting species.

@@

    %\subsection*{3 Global scale network properties are known to affect stability}
    The global structural properties of ecological networks have implications for their functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002d, Eklof2006}. For example, more highly connected networks are more resistant to secondary extinctions \citep{Dunne2002d, Eklof2006} and a modular organization, where some groups of species are tightly interconnected with limited connections between groups, can limit the propagation of disturbances \citep{}.
    %\subsection*{4 meso-scale nw propoerties also affect stability}
    Structural properties on the local scale, i.e. properties of single species, are also well-studied and there is consensus that species with high trophic levels, few prey (low in-degree), and low abundances are generally at higher risks of secondary extinction \citep{binzer2011susceptibility}. However, in-degree and trophic level are both relatively coarse descriptions of a species' role within its food-web~\citep{Cirtwill2018FoodWebs}. 
    For example, specialists may not necessarily have a high risk of extinction if it consumes a basal resource which is at low risk of being affected of disturbance. On the other hand, a consumer with low trophic level may still be at high risk of extinction if it consumes preys which possess high risk of being affected by disturbance.

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. Importantly, these motif roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions. 
    This means that two species with the same in-degrees and trophic levels can still have quite different motif roles if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. This extended description of species structural roles may give additional information of extinction risk. 
    At the network level, the frequencies of different motifs is associated with community stability \citep{prill2005dynamic, bascompte2005simple}.
    Some motifs, such as the three-species chain, are likely to retain all species within them when modelled in isolation~\citep{Borrelli2015a} and over-represented in empirical food webs~\citep{Stouffer2007}.
    This suggests that some motifs may increase network stability by reducing the probability that species participating in them will go extinct.
    The plant community has been shown to be particularly relevant for how the motifs are distributed within a network. For example, a recent study in experimental grasslands showed that changes in the plant community, i.e. richness, induced shifts in the relative frequency of motifs in the food webs studied \cite{giling2019plant}.

    %\subsection{6. only 4 (stable) motifs can appear in Bayesian networks, so how do we expect them to affect stability in a bottom-up context?}
    Of the 13 three-species motifs that can occur in empirical food webs there are four motifs that are over-represented in empirical networks \citep{Stouffer2007} and contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}: the three-species chain, omnivory, apparent competition and direct competition motifs.
    These four motifs are also the only ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. Because of this restriction, we only consider these four particularly important motifs.

       %\subsection*{Here we test...}
    Here we are interested in how disturbances at the basal level in food webs propagate up the network and whether a consumer's extinction risk depends on 1) the motif profile of the full food web and 2) the motifs in which a focal species participates. In particular, we want to understand whether the information provided by motifs can be explained by the classical global and local food web properties such as network connectance and species trophic level, or whether motifs provide unique insight into bottom-up extinction cascades.

@@

            We are interested in whether having a greater proportion of a species' role be made up of a certain motif (e.g., direct competition) might increase persistence.
            % (see \emph{Appendix SXX} for relationships between persistence and counts of motifs).
            To test this, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two.
            We centered and scaled the predictors before fitting the LMMs.
            The models also included a random intercept for the interaction between species richness and connectance (equations.
            All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

@@

            We were also interested in whether these general trends were consistent across networks with different global properties and different levels of disturbance. 
            To test this, we performed a linear regression of persistence against motif participation within each simulated network, for each level of disturbance to basal resources.
            These regressions were fit using the R~\citep{R} base function `lm'.
            We then analyse the distribution of slopes for the effect of motif participation across levels of disturbance, network size, and connectance.

@@

            Species' motif roles are often related to their degrees and trophic levels.
            To understand how this may have affected the results above, we first fit two linear mixed effects regressions relating persistence to degree or trophic level, $basal_p$, and their interaction.
            As above, we centered and scaled the predictors.
            We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks. 
            These regressions provide context about how persistence varies with simpler measures of species' roles within networks.

@@

            A further complication when trying to understand whether relationships between motifs and probability of extinction depend upon underlying relationships between degree or trophic level and extinction 
            We also fit linear mixed effect models relating the count or proportion of each motif to degree or trophic level and a random effect of the interaction between network size and connectance using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.

        \subsubsection*{How does persistence vary with motif positions?}

            As one more level of detail, we can consider the frequency with which a species appears in each \emph{position} in each motif. 
            Different positions are associated with different numbers of predators and prey, making the similarities and differences between motifs and degree more explicit.
            The full set of these positions, across all motifs, is the species' \emph{motif role}~\citep{Stouffer2010,Cirtwill2017}.
            As with motif participation, we normalized motif roles by dividing the number of times a species appeared in each position by the total number of times it appeared in all positions~\citep{Baker2015,Cirtwill2017}.

            [[Could add a PERMANOVA to show that species' roles overal affect their persistence. Haven't done so yet.]]

            To test whether participating in different motif positions was related to a species' probability of persistence, we fit similar models to those for motif participation.
            For each position, we fit a linear mixed effect model relating persistence to the frequency of the position in a species' motif role, disturbance, and their interaction.
            To account for differences in persistence across networks of different size and connectance, we included a random effect of the interaction between size and connectance.
            All regressions were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.

        \subsubsection*{How do motif roles vary with other network properties?}

            As with network motif profiles and motif participation, species' motif roles may vary with network size and connectance and with the species' in degree and trophic level.
            To identify these relationships, we fit two linear regressions for each motif position.
            The first related the frequency of the position in species' roles to network size, connectance, and their interaction.
            The second related the frequency of the position to in-degree, STL, and their interaction.
            All regressions were fit using the R~\citep{R} base function `lm'.

@@

    \subsection*{Persistence did not vary strongly with global network structure}

         Network size and connectance affected persistence, but to a much smaller extent than increasing disturbance to basal resources (Figure S1, Table S1, \emph{Appendix SA}). Persistences generally decrease when networks are more connected, but the effect of network size varies depending on level of disturbance. At low disturbance, small and highly-connected webs show the lowest persistence, whereas at high disturbance, large and highly connected webs show the lowest persistence (Fig. S2, \emph{Appendix SA}).

    \subsection*{Network motif profiles varied with global network properties}

        A PERMANOVA test suggests that a network's normalized motif profile as a whole (i.e., proportions of the total made up by each motif) is related to the network's size, connectance, and the interaction between them, but unequal variation in motif profiles across global network properties may, however, mean that these PERMANOVA results are false positives (Fig. S2, \emph{Appendix SB}).
        Considering each motif separately, the proportions of apparent competition and three-species chains in a network's motif profile tended to decrease with increasing connectance while the proportion of omnivory increased with increasing connectance (Table S2, \emph{Appendix SB}). 
        Network size and the interaction between network size and connectance did not strongly affect the frequencies of most motifs (Fig.~\ref{motif_proportion_lms}). The proportion of direct competition, however, decreased with increasing connectance in small networks but showed very little change in large webs.

@@

    \subsection*{Persistence decreases with greater omnivory in a network}

        The motif profile of a network was related to the mean persistence of consumers ($F_{1,2998}$=571, $p$\textless0.001). Mean persistence was higher in networks with higher proportions of the chain, direct competition, and apparent competition motifs (Table S3, \emph{Appendix SC}). Conversely, mean persistence is lower in networks with higher proportions of the omnivory motif. These trends hold for all levels of disturbance (Table S3, \emph{Appendix SC}). Dispersion of motif profiles varies between networks with different levels of persistence ($F_{134,2865}$=2.19$\times10^{24}$, $p$=\textless0.001). For example, networks with more variable motif profiles have higher persistence ($\beta$=0.183, $p$\textless0.001).

@@

    \subsection*{Species persistence varies with motif participation}

        % See \emph{Appendix SY} for relationships between persistence and non-normalized motif participation.

        \subsubsection*{Effects of motif participation depend on disturbance}

            At low levels of disturbance, species' persistences increased with increasing proportions of the omnivory and direct competition motifs. Conversely, persistence decreased with as the  proportion of the apparent competition motif increased. Persistence was not significantly related to the proportion of the three-species chain (Figure~\ref{fig:prop_lmer_all}).

            At high levels of disturbance, persistence decreased as the proportion of the three-species chain increased, but increased as the proportion of all other motifs increased (Figure~\ref{fig:prop_lmer_all}).
            % This means that increased disturbance strengthened the effects of the three-species chain motif and changes direction of the effect from the apparent competition motif  (Table S4, \emph{Appendix SD}).
            The effect of omnivory on persistence was  positive but weaker at higher levels of disturbance, and the effect of direct competition was largely unaffected by disturbance level.

@@

        \subsubsection*{Effects of motif participation are more consistent in highly-connected networks}

            By fitting linear regressions of persistence against motif proportions for the species for each network separately, we can observe how the relationships between motifs and persistence change with the level of disturbance and global network structure. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here. 
            Networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). 
            However, there was an interaction between disturbance and connectance for the direct competition and omnivory motifs.

            At low and medium connectances with a high proportion of the direct competition motif, there is little change in the proportion of positive slopes (\textgreater80\%) for the different levels of disturbance. 
            At high connectances, however, the proportion of positive slopes is lower overall and decreases with increasing disturbance. 
            This means that at low levels of connectance, an increased proportion of the direct competition motif increases persistence in most networks, while the opposite is true with high connectance. 
            The trend is reversed for the omnivory motif.

        \begin{figure}[h!]
            \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_C_allS.pdf}
            \caption{Density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all sizes - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. The slopes are derived from standard linear regression. Columns show the result for various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (left) to $\pi_{disturbed} = 0.5$ (right). Rows show various levels of connectance. The dotted, vertical line indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_C}
        \end{figure}

    \subsection*{Persistence increases with degree but decreases with STL}

        Persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix SF}). Increased trophic level is associated with decreasing persistence, although the rate of this decline is slower at higher levels of disturbance (Table S5, \emph{Appendix SF}).
        The interaction between in-degree and disturbance is even stronger, such as having a large number of prey is associated with higher persistence in the no disturbance scenario, but lower persistence at high disturbance.

@@

    \subsection*{Simpler measures do not fully explain effects of motifs}

        Motif participation varied with web size, connectance, and their interaction (Fig. S4, Table S6; \emph{Appendix SG}).
        Species in all networks appeared most frequently in the apparent competition motif, but the frequency of the apparent competition motif decreased with increasing network size and connectance.
        The frequency of the omnivory motif increased sharply with increasing connectance, while the frequency of the direct competition motif increased slightly with increasing network size.

        Species' motif roles also vary with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tend to have higher proportions of their motif role made up by omnivory. Proportions of the other three motifs tend to decline with increasing degree.
        Species with higher trophic levels tend to have higher proportions of apparent competition and three-species chains.

@@

    \subsection*{Persistence varies with a species' motif role}

        A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table SX, \emph{Appendix SH}).
        In the omnivory motif, persistence increased with the frequency of the bottom position (two predators, no prey within the motif) across all levels of disturbance but most steeply at high levels of disturbance (Fig.~\ref{figs:persistence_position_motifs}).
        Persistence increased slightly with the frequency of the middle position (one prey, one predator) at low levels of disturbance but showed little relationship at higher levels of disturbance.
        Persistence increased with the frequency of the top position (two prey, no predators within the motif) at low disturbance but decreased at higher levels of disturbance.

        In the three-species chain, persistence decreased as the frequency of the top position increased at all levels of disturbance.
        Conversely, persistence increased with increasing frequency of the middle position at all levels of disturbance.
        Persistence decreased slightly with increasing frequency of the bottom position at low levels of disturbance but increased at high levels of disturbance.

        In the apparent competition motif, persistence decreased slightly with increasing frequencies of the bottom position (one predator, no prey within the motif) at low disturbance but increased slightly at high disturbance.
        Persistence increased with increasing frequency of the top position (two prey, no predator within the motif) at low disturbance but decreased at high disturbance.

        In the direct competition motif, persistence varied little with the frequency of the bottom position (two predators, no prey within the motif) at low disturbance but increased with the frequency of this position at high levels of disturbance.
        Persistence increased slightly with the frequency of the top position (one prey, no predators within the motif) at low levels of disturbance and decreased slightly at high levels of disturbance.

    \begin{figure}[h!]
        \includegraphics[width=.95\textwidth]{figures/persistence_positions_Omnivory.eps}
        \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the omnivory motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
        \label{figs:persistence_position_motifs}
        \end{figure}

        The frequencies of motif positions in a species' role varied with network properties (\emph{Appendix SI}).
        Motif roles had stronger relationships with degree and STL than with network size or connectance.

@@

%Somewhere early in discussion (needs tidying up):
%Apparent comp and direct comp have no indirect effects in BN, so we should be able to explain them with simpler properties.
%AC is strongly negatively correlated with in-degree. Higher in degree is associated with both high persistence at low disturbance (Fig. 7A) and low participation in AC (Fig. 7C). Low participation in AC is in turn associated with high persistence at low disturbance (Fig. 4). Note that trends for degree and AC reverse at high disturbance in the same way!
%DC is not strongly correlated with degree but does decrease strongly at high trophic levels. High trophic level and low DC are both associated with low persistence at all levels of disturbance, making it likely that the effects we observe for DC are trophic level in disguise.
%Three-species chain and omnivory do have indirect effects in a BN setting, and they can't be fully explained by degree or trophic level (since the result doesn't follow the result for degree). We can see this because despite the strong correlations with degree and trophic level these motifs don't show the same patterns with persistence. We therefore focus on them here.

>>>>

% Maybe Proc B - approached Alyssa re: other motifs paper, flexible length limit (but pay extra over X pages).

@@

    - can we condense omnivory, in-degree any further?
    - - at least omnivory
    - can we shorten any more? Nice to save some money?
    - glossary
    - refs
    - ending

    - repeat network size analyses with \# basal resources - maybe adjust Fig. 1 to reflect n(basal) and C instead of C:S?
    - Fig. 4 not much in discussion, could be a candidate for the SI? Or add more discussion about apparent competition, etc. in discussion sections per motif

\clearpage
\section*{Introduction}

    %\subsection*{1 Why are bottom-up effects important for understanding stability?}
     The plant community is the foundation upon which the myriad of species in food webs depend for their survival \citep{}. Disturbances in the plant community have been shown to affect important ecosystem properties such as primary \citep{} and secondary production \citep{borer2012plant}, soil respiration and carbon cycling \citep{chen2019plant} and consumer diversity \citep{scherber2010bottom, Baiser2016}. The stability of the plant community is crucial for sustaining a healthy energy flow in the ecosystem as a whole \citep{Rosenblatt2016} and the structure of the plant community promotes stability across all different levels of ecosystem organization \citep{proulx2010diversity,scherber2010bottom}. Loss of diversity at the basal level typically causes declines in both abundance and richness of all types of consumers, e.g. herbivores, predators, and parasitoids \citep{scherber2010bottom}.
     %This has been shown in both theoretical work~\citep{} and empirical experiments ~\citep{} and results from the vertical propagation of disturbances through the network of interacting species.

@@

    %\subsection*{3 Global scale network properties are known to affect stability}
    The global structural properties of ecological networks have implications for their functioning \citep{Petchey2002}, stability \citep{Allesina2012} and robustness to disturbances \citep{Dunne2002d, Eklof2006}. For example, more highly-connected networks are more resistant to secondary extinctions \citep{Dunne2002d, Eklof2006} and a modular organization, where some groups of species are tightly interconnected with limited connections between groups, can limit the propagation of disturbances \citep{}.
    %\subsection*{4 meso-scale nw propoerties also affect stability}
    Structural properties on the local scale, i.e. properties of single species, are also well-studied and there is consensus that species with high trophic levels, few prey (low in-degree), and low abundances are generally at higher risks of secondary extinction \citep{binzer2011susceptibility}. However, in-degree and trophic level are both relatively coarse descriptions of a species' role within its food-web~\citep{Cirtwill2018FoodWebs}. 
    For example, specialists may not necessarily have a high risk of extinction if it consumes a basal resource which is at low risk of being affected of disturbance. On the other hand, a consumer with low trophic level may still be at high risk of extinction if it consumes preys which themselves possess high risk of being affected by disturbance.

    %\subsection*{5 Motif participation may similarly affect species' particular extinction risk}
    One way to incorporate more nuance into descriptions of how species fit into their communities is to define species' roles based on their participation in different motifs -- unique combinations of small numbers of interacting species~\citep{Stouffer2007,Stouffer2012}. Importantly, these motif roles capture information about species' direct and indirect interactions, incorporating larger-scale network structure into species-level descriptions. 
    This means that two species with the same in-degrees and trophic levels can still participate in different sets of motifs if, for example, one interacts with specialists and the other interacts with generalists~\citep{Cirtwill2018FoodWebs}. 
    This extended description of species structural roles may give additional information of extinction risk or synthesize the information provided by multiple other measures of a species' place in its community.
    At the network level, the frequencies of different motifs are associated with community stability \citep{prill2005dynamic, bascompte2005simple} and with changes to the composition of plant communities~\cite{giling2019plant}. 
    We therefore expect that the motif structure of a network and species' participation in that structure should be related to risks due to disturbance to basal resources.

    %\subsection{6. only 4 (stable) motifs can appear in Bayesian networks}
    Of the 13 three-species motifs that can occur in empirical food webs there are four motifs that are over-represented in empirical networks and are thought to contribute to stability \citep{Stouffer2007, Borrelli2015a, giling2019plant}: the three-species chain, omnivory, apparent competition and direct competition motifs.
    These four motifs are also the ones which can occur in the acyclic networks required for Bayesian network simulations~\citep{Eklof2013}. Therefore, we focus on these four particularly important motifs.

       %\subsection*{Here we test...}
    Here we are interested in how disturbances at the basal level in food webs propagate up the network and whether a consumer's extinction risk depends on 1) the motif profile of the full food web and 2) the motifs in which a focal species participates. In particular, we want to understand whether the information provided by motifs is related to classical global and local food web properties such as network connectance and species trophic level, or whether motifs provide unique insight into bottom-up extinction cascades.

@@

            We are interested in whether participating more often in a certain motif (e.g., direct competition) might increase persistence.
            % (see \emph{Appendix SXX} for relationships between persistence and counts of motifs).
            To test this, we fit four linear mixed-effect models (LMMs; one per motif included in the Bayesian networks).
            These models included the proportion of the species' role made up by the focal motif, the level of disturbance to basal species, and the interaction between the two.
            We centered and scaled the predictors before fitting the LMMs.
            The models also included a random intercept for the interaction between species richness and connectance (equations.
            All models were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We calculated marginal (fixed effects only) and conditional (fixed and random effects) $R^2$ using the R~\citep{R} function `r.squaredGLMM' from the package \emph{MuMIn}~\citep{MuMIn}.

@@

            We were also interested in whether these general trends were consistent across networks with different global properties and different levels of disturbance. 
            To test this, we performed a linear regression of persistence against motif participation within each simulated network, for each level of disturbance to basal resources.
            These regressions were fit using the R~\citep{R} base function `lm'.
            We then analyse the distribution of slopes for the effect of motif participation across levels of disturbance, network size, and connectance. 
            We performed similar analyses to test whether appearing in different positions in each motif (i.e., the motif role) is related to a species' persistence (\emph{Appendix SH}).

@@

            A species' motif participation is likely related to its degrees and trophic level since all three are measures of network position.
            To understand how this may have affected the results above, we first fit two linear mixed-effects regressions relating persistence to degree or trophic level, $\pi_{disturbed}$, and their interaction.
            As above, we centered and scaled the predictors.
            We also included a random effect of the interaction between network size and connectance in order to account for the differences in persistence between different networks. 
            These regressions provide context about how persistence varies with simpler measures of species' roles within networks.

@@

            A further complication when trying to understand whether relationships between motifs and probability of extinction depend upon underlying relationships between degree or trophic level and extinction 
            We also fit linear mixed-effect models relating the count or proportion of each motif to degree or trophic level and a random effect of the interaction between network size and connectance using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.
            We performed similar analyses to test whether a species' motif role is related it in-degree or trophic level (\emph{Appendix SI}).

        % \subsubsection*{How does persistence vary with motif positions?}

        %     As one more level of detail, we can consider the frequency with which a species appears in each \emph{position} in each motif. 
        %     Different positions are associated with different numbers of predators and prey, making the similarities and differences between motifs and degree more explicit.
        %     The full set of these positions, across all motifs, is the species' \emph{motif role}~\citep{Stouffer2010,Cirtwill2017}.
        %     As with motif participation, we normalized motif roles by dividing the number of times a species appeared in each position by the total number of times it appeared in all positions~\citep{Baker2015,Cirtwill2017}.

        %     [[Could add a PERMANOVA to show that species' roles overal affect their persistence. Haven't done so yet.]]

        %     To test whether participating in different motif positions was related to a species' probability of persistence, we fit similar models to those for motif participation.
        %     For each position, we fit a linear mixed-effect model relating persistence to the frequency of the position in a species' motif role, disturbance, and their interaction.
        %     To account for differences in persistence across networks of different size and connectance, we included a random effect of the interaction between size and connectance.
        %     All regressions were fit using the R~\citep{R} function `lmer' from the package \emph{lmerTest}~\citep{lmerTest}.

        % \subsubsection*{How do motif roles vary with other network properties?}

        %     As with network motif profiles and motif participation, species' motif roles may vary with network size and connectance and with the species' in degree and trophic level.
        %     To identify these relationships, we fit two linear regressions for each motif position.
        %     The first related the frequency of the position in species' roles to network size, connectance, and their interaction.
        %     The second related the frequency of the position to in-degree, STL, and their interaction.
        %     All regressions were fit using the R~\citep{R} base function `lm'.

@@

    We are interested in i) how the distribution of motifs in a network (i.e., the network's \emph{motif profile}) affect network persistence, and ii) how a species motif participation affects its persistence. We analyse this for different levels of disturbance to the basal level. To provide context, we also investigate relationships between persistence and classic global (size and connectance) and local (in-degree and trophic level) network properties.
    % The results are organized as follows: 1) How network persistence varies with network motif distribution. 2) How network persistence varies with global structural properties. 3) How network motif distribution varies with global structural properties. 4) How species probability of persistence varies with its motif profile. 5) How species persistence varies with its characteristics such as TL and in-degree. 6) How a species motif profile depend on its characteristics. 7) How a species persistence varies with its motif role (position in respective motif).

    \subsection*{Mean persistence, disturbance, and global network properties}

        The mean probability of persistence for all consumers in a network naturally decreased with increasing probability of extinction of basal species. These effects were clear across the ranges of network size and connectance we considered. 
        Network size and connectance also affect network mean persistence, but to a much smaller extent than increasing disturbance (Figure S1, Table S1, \emph{Appendix SA}). Persistences generally decrease when networks are more connected, but the effect of network size varies depending on level of disturbance. At low disturbance, small and highly-connected webs show the lowest persistence, whereas at high disturbance, large and highly connected webs show the lowest persistence (Fig. S2, \emph{Appendix SA}).

    \subsection*{Network motif profiles and global network properties}

        Connectance was significantly related to the frequencies of different motifs in a network.
        Network size and the interaction between network size and connectance did not strongly affect the frequencies of most motifs (Fig.~\ref{motif_proportion_lms}).
        Considering each motif separately, the proportions of apparent competition and three-species chains in a network's motif profile tended to decrease with increasing connectance while the proportion of omnivory increased with increasing connectance (Table S2, \emph{Appendix SB}).

@@

    \subsection*{Network motif profiles and network persistence}

        The motif profile of a network is related to the average persistence of consumers ($F_{1,2998}$=571, $p$\textless0.001). Average persistence is higher in networks with higher proportions of the chain, direct competition, and apparent competition motifs (Table S3, \emph{Appendix SC}). Conversely, average persistence is lower in networks with higher proportions of the omnivory motif. These trends hold for all levels of disturbance (Fig.~\ref{motif_profile_persistence}, Table S3), \emph{Appendix SC}).% Network persistence varies between networks with different dispersion of motif profiles ($F_{134,2865}$=2.19$\times10^{24}$, $p$=\textless0.001). For example, networks with more variable motif profiles have higher persistence ($\beta$=0.183, $p$\textless0.001).

@@

    \subsection*{Species motif participation and species persistence}

       The effects of motif participation on a species' probability of persistence depends on the level of disturbance. When disturbance is low, species' persistence increases with increasing proportions of the omnivory and direct competition motifs. Conversely, a species' persistence decreases with an increasing proportion of the apparent competition motif. Persistence is not significantly related to the proportion of the three-species chain motif (Figure~\ref{fig:prop_lmer_all}).

        At high levels of disturbance, a species' persistence decreases as the proportion of the three-species chain increases, but persistence increases as the proportion of any other motif increases (Figure~\ref{fig:prop_lmer_all}). This means that increased disturbance strengthens the effects of the three-species chain motif and changes the direction of the effect of the apparent competition motif  (Table S4, \emph{Appendix SD}). The effect of omnivory on persistence is still positive but weaker at higher levels of disturbance, and the relationship between persistence and direct competition was largely unaffected by disturbance.

@@

        \subsubsection*{Consistency in effects of motif participation on species persistence}

            % In order to evaluate how consistent the relationships between a species persistence and motif participation is, and how it varies with  level of disturbance and global network structure, we fit linear regressions of species persistence against motif proportions for the species for each network separately. We then analyze the density distribution of the values of the slopes. Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}); we therefore focus on connectance here.

            When fitting regressions of persistence against motif participation for each network separately, networks with high connectance generally have more consistent relationships between persistence and motif participation, i.e., sharper peaks of the density distributions (Fig.~\ref{fig:density_prop_C}). However, there is an interaction between disturbance and connectance for the direct competition and omnivory motifs; At low and medium connectances there is little change in the proportion of positive slopes when the proportion of  direct competition motif is high, independent of level of disturbance. At high connectances, however, the proportion of positive slopes is lower and also decreases with increasing disturbance. For the omnivory motif the trend is reversed. Additionally, the omnivory motif stands out having a wider curve when network connectance is low. 
            Network size had a much smaller effect than connectance (Fig. S3; \emph{Appendix SE}).

        \begin{figure}[h!]
            \centering
            \includegraphics[width=\textwidth]{figures/prop_dens_bp_vs_C_allS.pdf}
            \caption{Density (y-axis) of slopes (x-axis) of persistence against proportion of different motifs for all simulated webs of all sizes - a visualization of how an increased proportion of each motif (different colored lines) affects persistence of consumer species. Columns show the result for various disturbances on the basal level, from $\pi_{disturbed} = 0.1$ (left) to $\pi_{disturbed} = 0.5$ (right). Rows show various levels of connectance. The dotted, vertical line indicate zero on the x-axis. A negative slope value reflects a negative relationship between increased participation in a motif and persistence, while a positive slope value reflects a positive relationship - an increased proportion of a specific motif increase persistence. The fraction of replicates with a slope greater than zero are stated in numbers in each sub-plot, the color corresponding to each motif (legend). }
            \label{fig:density_prop_C}
        \end{figure}

    \subsection*{Species characteristics and persistence}

        A species persistence varies with both species in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}; \emph{Appendix SF}). Increased trophic level is generally associated with decreasing species persistence, although the rate of decline is slower at high disturbance (Table S5, \emph{Appendix SF}).
        In-degree shows an even stronger interaction with disturbance; having more prey species is associated with higher persistence without disturbance, but lower persistence at high levels of disturbance.

@@

    \subsection*{Motif participation, network properties, and species characteristics}

       Species' motif participation varied depending on the properties of the network; size, connectance, and their interaction (Fig. S4, Table S6; \emph{Appendix SG}).
        Species appeared most often in the apparent competition motif in all types of networks, but the frequency decreased with increasing network size and connectance.
        The frequency with which species participated in the omnivory motif increased sharply with increasing connectance, while the frequency of the direct competition motif increased slightly with increasing network size.

        Species' motif participation also varied with their in-degree and trophic level (Fig.~\ref{fig:motifs_vs_TL_and_deg}).
        Species with higher in-degree tend to have higher frequencies of participation in the omnivory motif. Proportions of the other three motifs tend to decline with increasing degree.
        Species with higher trophic levels tend to have higher proportions of apparent competition and three-species chains.

@@

    % \subsection*{Persistence and species' positioning in motifs}

    %     A species probability of persisting is significantly related to the frequency of all positions, disturbance, and the interaction between frequency of each position and disturbance (Table SX, \emph{Appendix SH}).
    %     In the omnivory motif, persistence increased with the frequency of the bottom position (two predators, no prey within the motif) across all levels of disturbance but most steeply at high levels of disturbance (Fig.~\ref{figs:persistence_position_motifs}).
    %     Persistence increased slightly with the frequency of the middle position (one prey, one predator) at low levels of disturbance but showed little relationship at higher levels of disturbance.
    %     Persistence increased with the frequency of the top position (two prey, no predators within the motif) at low disturbance but decreased at higher levels of disturbance.

    %     In the three-species chain, persistence decreased as the frequency of the top position increased at all levels of disturbance.
    %     Conversely, persistence increased with increasing frequency of the middle position at all levels of disturbance.
    %     Persistence decreased slightly with increasing frequency of the bottom position at low levels of disturbance but increased at high levels of disturbance.

    %     In the apparent competition motif, persistence decreased slightly with increasing frequencies of the bottom position (one predator, no prey within the motif) at low disturbance but increased slightly at high disturbance.
    %     Persistence increased with increasing frequency of the top position (two prey, no predator within the motif) at low disturbance but decreased at high disturbance.

    %     In the direct competition motif, persistence varied little with the frequency of the bottom position (two predators, no prey within the motif) at low disturbance but increased with the frequency of this position at high levels of disturbance.
    %     Persistence increased slightly with the frequency of the top position (one prey, no predators within the motif) at low levels of disturbance and decreased slightly at high levels of disturbance.

    % \begin{figure}[h!]
    %     \includegraphics[width=.95\textwidth]{figures/persistence_positions_Omnivory.eps}
    %     \caption{A species' probability of persistence varied with the frequency of different motif positions in its role. Here we show the relationships between persistence and the proportions of the three positions in the omnivory motif at three levels of probability of extinction for basal resources (p=0.1, 0.26, and 0.42), indicated by line colour. Line style indicates the motif  position. }
    %     \label{figs:persistence_position_motifs}
    %     \end{figure}

    %     The frequencies of motif positions in a species' role varied with network properties (\emph{Appendix SI}).Motif roles had stronger relationships with degree and STL than with network size or connectance.

@@

%Somewhere early in discussion (needs tidying up):
%Apparent comp and direct comp have no indirect effects in BN, so we should be able to explain them with simpler properties.

\section*{Discussion}

% Focus on what additional information is revealed from analyzing motif compared to global network properties or species characteristics.
% Connect to traits -- Cirtwill & Eklöf, 2018, Ecol Lett. 
% Bayesian network approach - what extra information does motifs provide when we are interested in bottom- up effects. 
% * Difference between 3-sp chain and omnivory, the importance of the rescue prey in omnivory (in dynamical sims omni was mostly bad, 3-sp chain good). Omni link destabilizing if strong, stabilizing if weak (other studies). In BN just a link as important as other links
% * Omnivory in fig 5 - high connectance, always good, not as good with lower levels of C
% * Extra prey in omnivory changes from being a security to a liability (with increasing disturbance?)
% * ADD REFS - high disturbance scenario is analogous to decreasing plant diversity
